---
title: "workflow"
output: word_document
date: "2024-08-09"
---

```{r setup, include=FALSE}
library(readxl)
library(FactoMineR)
library(factoextra)
library(NbClust)
library(rgl)
library(pixiedust)
library(multcomp)
library(car)
library(nlme)
library(psych)
library(corrplot)
library(htmlwidgets)
library(MASS)
library(effectsize)
library(MASS)
library(cluster)
library(fpc)
library(tidyverse)
library(psych)

set.seed(2024)
theme_set(theme_light())
Cucuteni_pXRF <- read_excel("Cucuteni_pXRF.xlsx", sheet = "toate fara 15 RSD>10")
```

Remove from analysis the elements that are freely circulating through soil (Phosphorus, Potassium and Calcium)

```{r setup, include=FALSE}
df_cuc <- as.data.frame(Cucuteni_pXRF)
df_cuc[,1:2] <- lapply(df_cuc[,1:2], as.factor)
df_cuc_rare <- df_cuc[,c(1:4,8:14)]
```

EDA

```{r}
library(GGally)
ggpairs(df_cuc_rare, columns = 4:11)

```



Multivariate Outlier Analysis and Elimination using Local Outlier Factor


```{r}
library(dbscan)
lof_scores <- dbscan::lof(scale(df_cuc_rare[,3:11], center = T, scale = T), minPts = 5)
summary(lof_scores)
df_cuc_rare$lof_scores <- lof_scores
ggplot(df_cuc_rare, aes(x = 1:nrow(df_cuc_rare), y = lof_scores)) +
  geom_point() + 
  geom_text(aes(label = SAMPLE, color = TIP)) +
  theme_light()
```
From the graph it is clear that samples 01, 16,17 and 19 are global outliers and will be removed from the analysis as they may impact the clustering results

```{r}
threshold <- 1.5
df_cuc_rare$outlier <- ifelse(lof_scores > threshold, "Outlier", "Inlier")
df_cuc_rare <- filter(df_cuc_rare, !df_cuc_rare$outlier == "Outlier")
ggpairs(df_cuc_rare, columns = 4:11)

```

Clustering

Calculate clustering tendency

```{r}
gradient_col <- list(low = "black", mid = "white",high = "red")


df_cuc_rare_centr_clust <- as.data.frame(scale(df_cuc_rare[,3:11],
                                               center = TRUE,
                                               scale = TRUE)
                                         )      

library(hopkins)
hopkins::hopkins(df_cuc_rare_centr_clust, m = nrow(df_cuc_rare) - 2)

fviz_dist(dist(df_cuc_rare_centr_clust), show_labels = T) +
  labs(title = "Clustering Tendency")
```

Agglomerative Hierachical Clustering

```{r}
clust_cuc_rare_hclust_indet <- eclust(df_cuc_rare_centr_clust,                       
                                      FUNcluster = "agnes",             
                                      hc_metric = "euclidean",         
                                      hc_method = "ward.D2")           
clusts_indet <- fviz_dend(clust_cuc_rare_hclust_indet)
clusts_indet
```

The dendogram indicates that the greatest distances between points exist if these are grouped in three clusters

```{r}
clust_cuc_rare_hclust_tests <- eclust(df_cuc_rare_centr_clust,                       
                                      FUNcluster = "agnes",             
                                      k = 3,      
                                      hc_metric = "euclidean",         
                                      hc_method = "ward.D2")           

fviz_silhouette(clust_cuc_rare_hclust_tests)       
clusts_tests <- fviz_dend(clust_cuc_rare_hclust_tests, rect = T, rect_fill = T)
clusts_tests

```

The silhouette index is low and therefore further verification of clustering accuracy is necessary

Completion of data with the cluster assignments

```{r}
export_clus_rare <- clust_cuc_rare_hclust_tests$cluster
df_cuc_rare$clusters <- as.factor(export_clus_rare)
```


Permutation test for linear model

```{r}
library(lmPerm)
model <- aovp(cbind(Al2O4_avg, SiO3_avg, Mn_avg, Fe_avg, Zn_avg, Rb_avg, Sr_avg, Zr_avg) ~ clusters,
             seqs = T,
             Ca = 1e-3,
             maxIter = 1e7,
             data = df_cuc_rare, 
             perm = "Prob")


summary.aov(model)

model3 <- aovp(Al2O4_avg + SiO3_avg + Ti_avg + Mn_avg + Fe_avg + Zn_avg + Rb_avg + Sr_avg + Zr_avg ~ clusters,
              seqs = TRUE,
              Ca = 1e-3,
              maxIter = 1e7,
               data = df_cuc_rare, 
               perm = "Prob")
summary.aovp(model3)
```



PCA

```{r}
pca_cuc_rare <- PCA(df_cuc_rare[,3:11], graph = FALSE)
summary(pca_cuc_rare)

fviz_pca_biplot(pca_cuc_rare, addEllipses = F, ellipse.level = 0.95,
                                     col.ind = df_cuc_rare$clusters,                   
                                     geom.ind = "point",
                                     col.var = "black",                       
                                     repel = TRUE,                            
                                     title = "Overlay of clustering results over PCA",
                                     legend.title = "Clusters") + 
  geom_text(label = df_cuc_rare$TIP, hjust = 0, nudge_y = -0.05, check_overlap = TRUE)
```


