---
title: "workflow"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
date: "2024-08-09"
---

```{r setup, include=FALSE}
library(readxl)
library(FactoMineR)
library(factoextra)
library(NbClust)
library(rgl)
library(pixiedust)
library(multcomp)
library(car)
library(nlme)
library(psych)
library(corrplot)
library(htmlwidgets)
library(MASS)
library(effectsize)
library(MASS)
library(cluster)
library(fpc)
library(tidyverse)
library(psych)
library(janitor)
library(patchwork)
library(viridis)

set.seed(2024)
theme_set(theme_light())

Cucuteni_pXRF <- read_excel("C:/Users/georg/OneDrive/Proiecte_R/Proiect-R-Cucuteni/Cucuteni_pXRF.xlsx", sheet = "toate fara 15 RSD>10")
```

#Data Selection
##Removing variables


Remove from analysis the elements that are troublesome for quantitative analysis (Phosphorus per Hunt & Speakman 2015 JAS 53)

```{r}
df_cuc <- as.data.frame(Cucuteni_pXRF)
df_cuc[,1:2] <- lapply(df_cuc[,1:2], as.factor)
df_cuc <- df_cuc |> 
  unite("Proba_Tip", TIP:SAMPLE, sep = "_", remove = F) |> 
  column_to_rownames("Proba_Tip") |> 
  unite("Proba_Tip", TIP:SAMPLE, sep = "_", remove = F)
df_cuc_rare <- df_cuc[,c(1:5, 7:15)]

```

#EDA

```{r}
library(GGally)
ggpairs(df_cuc_rare, columns = 4:14)

```

Following EDA we will also eliminate Al due to its high correlation with Ti, Fe, and Zn, and K, due to its high correlation with Rubidium and Si. In this last case, we chose to eliminate K also due the possibility of contamination (Potassium per Frankel and Webb JAS 2012)

```{r}
df_cuc_rare <- df_cuc_rare[, -c(4,6)]
df_cuc_check_point <- df_cuc_rare
```


#Data Selection
##Removing individuals
###Multivariate Outlier Analysis and Elimination using Local Outlier Factor


```{r}
library(dbscan)
lof_scores <- dbscan::lof(scale(df_cuc_rare[,4:12], center = T, scale = T), minPts = 5)
summary(lof_scores)
df_cuc_rare$lof_scores <- lof_scores
ggplot(df_cuc_rare, aes(x = 1:nrow(df_cuc_rare), y = lof_scores)) +
  geom_point() + 
  geom_text(aes(label = SAMPLE, color = TIP)) +
  theme_light()
```
From the graph it is clear that samples 01, 16, 17, 18, 19 are global outliers and will be removed from the analysis as they may impact the clustering results

```{r}
threshold <- 1.3
df_cuc_rare$outlier <- ifelse(lof_scores > threshold, "Outlier", "Inlier")
df_cuc_rare_with_outliers <- df_cuc_rare
df_cuc_rare <- filter(df_cuc_rare, !df_cuc_rare$outlier == "Outlier")
ggpairs(df_cuc_rare, columns = 4:12)

```

#Clustering

##Calculate clustering tendency

```{r}
gradient_col <- list(low = "black", mid = "white",high = "red")


df_cuc_rare_centr_clust <- as.data.frame(scale(df_cuc_rare[,4:12],
                                               center = TRUE,
                                               scale = TRUE)
                                         )
library(hopkins)
hopkins::hopkins(df_cuc_rare_centr_clust, m = nrow(df_cuc_rare) - 2)

fviz_dist(dist(df_cuc_rare_centr_clust), show_labels = T) +
  labs(title = "Clustering Tendency")
```

##Agglomerative Hierachical Clustering

```{r}
clust_cuc_rare_hclust_indet <- eclust(df_cuc_rare_centr_clust,                       
                                      FUNcluster = "agnes",             
                                      hc_metric = "euclidean",         
                                      hc_method = "ward.D2")           
clusts_indet <- fviz_dend(clust_cuc_rare_hclust_indet)
clusts_indet

library(clValid)
distance_matrix <- distance(df_cuc_rare_centr_clust)
distance_matrix <- as.matrix(distance_matrix)
internal_validation <- clValid(distance_matrix, 2:7, 
                               clMethods = c("agnes"),
                               validation = c("stability"))
summary(internal_validation)

```

The dendrogram indicates that the greatest distances between grpups exist if these are grouped in three clusters

```{r}
clust_cuc_rare_hclust_tests <- eclust(df_cuc_rare_centr_clust,                       
                                      FUNcluster = "agnes",             
                                      k = 3,      
                                      hc_metric = "euclidean",         
                                      hc_method = "ward.D2")           

fviz_silhouette(clust_cuc_rare_hclust_tests)       
clusts_tests <- fviz_dend(clust_cuc_rare_hclust_tests, rect = T, rect_fill = T)
clusts_tests

```

The silhouette index is low and it indicates possible erroneous assignments in Cluster 1, therefore further verification of clustering accuracy is necessary

Completion of data with the cluster assignments

```{r}
export_clus_rare <- clust_cuc_rare_hclust_tests$cluster
df_cuc_rare$clusters <- as.factor(export_clus_rare)
```


#Permutation test for linear model

```{r}
library(lmPerm)
model <- aovp(cbind(SiO3_avg, Ca_avg, Ti_avg, Mn_avg, Fe_avg, Zn_avg, Rb_avg, Sr_avg, Zr_avg) ~ clusters,
             seqs = T,
             Ca = 1e-3,
             maxIter = 1e6,
             data = df_cuc_rare, 
             perm = "Prob")
summary.aovp(model)

model3 <- aovp(SiO3_avg + Ca_avg + Ti_avg + Mn_avg + Fe_avg + Zn_avg + Rb_avg + Sr_avg + Zr_avg ~ clusters,
              seqs = TRUE,
              Ca = 1e-3,
              maxIter = 1e6,
               data = df_cuc_rare, 
               perm = "Prob")
summary.aovp(model3)
```



#PCA with clusters overlay

```{r}
pca_cuc_rare <- PCA(df_cuc_rare[,4:12], graph = FALSE, scale.unit = T)
summary(pca_cuc_rare)

plot_pca <- fviz_pca_biplot(pca_cuc_rare, addEllipses = T, ellipse.level = 0.90,
                                     col.ind = df_cuc_rare$clusters,
                                     geom.ind = "point",
                                     col.var = "black",                       
                                     repel = TRUE,                            
                                     title = "Overlay of clustering results over PCA",
                                     legend.title = "Clusters") + 
  geom_text(label = df_cuc_rare$TIP, hjust = 0.5, nudge_y = -0.09, check_overlap = TRUE) +
  geom_text(label = df_cuc_rare$SAMPLE, hjust = 0.5, nudge_y = 0.09, nudge_x = -0.09, check_overlap = TRUE) +
  scale_colour_viridis_d()

plot_pca
```

######PCA - visualisations - eigen values
```{r}
eigen_values <- pca_cuc_rare$eig |> 
  as_tibble() |> 
  rownames_to_column() |> 
    janitor::clean_names() |> 
  arrange(desc(percentage_of_variance)) 


graph_eigen <- eigen_values |> 
    #top_n(4, percentage_of_variance) |>
  mutate(percentage_of_variance = round(percentage_of_variance, 2)) |> 
  ggplot(aes(rowname, percentage_of_variance)) +
  geom_col() +
  geom_text(aes(label = percentage_of_variance, vjust = 1.5,
              )) +
  xlab(label = "Principal Components") +
  ylab(label = "Percentage of Variance Explained") +
  labs(title = "Explanatory Power of the Identified Principal Components")
graph_eigen
```

######PCA - visualisations - variables
```{r}
contrib_var_selection <- pca_cuc_rare$var$contrib |> 
  as.data.frame() |> 
  rownames_to_column() |> 
  mutate_if(is.character, as.factor) 

graph_dim1_contrib <- 
  contrib_var_selection |> 
  mutate(rowname = fct_reorder(rowname, Dim.1)) |> 
  ggplot(aes(Dim.1, rowname)) +
  geom_col() +
  xlab(label = "Contribution of species to PC 1") +
  ylab(label = "")


graph_dim2_contrib <- 
  contrib_var_selection |> 
  mutate(rowname = fct_reorder(rowname, Dim.2)) |> 
  ggplot(aes(Dim.2, rowname)) +
  geom_col() +
  xlab(label = "Contribution of species to PC 2") +
  ylab(label = "")


graph_dim3_contrib <- 
  contrib_var_selection |> 
  mutate(rowname = fct_reorder(rowname, Dim.3)) |> 
  ggplot(aes(Dim.3, rowname)) +
  geom_col() +
  xlab(label = "Contribution of species to PC 3") +
  ylab(label = "")

cors_var <- pca_cuc_rare$var$cor |> 
  as.data.frame() |> 
  rownames_to_column() |> 
  mutate_if(is.character, as.factor) 

graph_dim1_cors <- 
  cors_var |> 
  mutate(rowname = fct_reorder(rowname, Dim.1, .desc = T)) |> 
  ggplot(aes(Dim.1, rowname)) +
  geom_col() +
  theme(axis.text = element_text(size = 9, face = "bold"),
        axis.title = element_text(face = "bold")
        ) + 
  xlab(label = "Correlation of species to the first principal component") +
  ylab(label = "") +
  inset_element(graph_dim1_contrib,
                left = 0.70,
                bottom = 0.5,
                right = 1,
                top = 1)

graph_dim2_cors <- 
  cors_var |> 
  mutate(rowname = fct_reorder(rowname, Dim.2, .desc = T)) |> 
  ggplot(aes(Dim.2, rowname)) +
  geom_col() +
  theme(axis.title = element_text(face = "bold"),
        axis.text = element_text(size = 9, face = "bold")
        ) + 
  xlab(label = "Correlation of species to the second principal component") +
  ylab(label = "") +
  inset_element(graph_dim2_contrib,
                left = 0.70,
                bottom = 0.5,
                right = 1,
                top = 1)


graph_dim3_cors <- 
  cors_var |> 
  mutate(rowname = fct_reorder(rowname, Dim.3, .desc = T)) |> 
  ggplot(aes(Dim.3, rowname)) +
  geom_col() +
  theme(axis.title = element_text(face = "bold"),
        axis.text = element_text(size = 9, face = "bold")
        ) + 
  xlab(label = "Correlation of species to the third principal component") +
  ylab(label = "") +
  inset_element(graph_dim3_contrib,
                left = 0.70,
                bottom = 0.45,
                right = 1,
                top = 1)
```

######PCA - visualisations - cases
```{r}
coord_ind <- 
  pca_cuc_rare$ind$coord |> 
  as.data.frame() |> 
  mutate_if(is.character, as.factor) |> 
  rownames_to_column() |> 
  as_tibble() |> 
  separate_wider_delim(rowname, delim = "_", 
                       names = c("Tip", "Proba", "Numar"),
                       cols_remove = F)

graph_ind_1 <- coord_ind |> 
  mutate(rowname = fct_reorder(rowname, Dim.1, .desc = T)) |> 
  ggplot(aes(Dim.1, rowname)) +
  geom_col(aes(colour = df_cuc_rare$clusters, fill = df_cuc_rare$clusters)) +
  theme(legend.position = "inside",
    legend.position.inside = c(1,1),
    legend.justification.inside = c(1,1),
    axis.text = element_text(size = 9, face = "bold"),
    axis.title = element_text(face = "bold")
    ) + 
  xlab(label =  "The sites' positioning in relation to the first principal component") + 
  ylab(label =  "Sites") +
  scale_fill_viridis(discrete = T) +
  scale_color_viridis_d()

graph_ind_2 <- coord_ind |> 
  mutate(rowname = fct_reorder(rowname, Dim.2, .desc = T)) |> 
  ggplot(aes(Dim.2, rowname)) +
  geom_col(aes(colour = df_cuc_rare$clusters, fill = df_cuc_rare$clusters)) +
  theme(legend.position = "inside",
    legend.position.inside = c(1,1),
    legend.justification.inside = c(1,1),
    axis.text = element_text(size = 9, face = "bold"),
    axis.title = element_text(face = "bold")
    ) + 
  xlab(label =  "The sites' positioning in relation to the second principal component") + 
  ylab(label =  "Sites") +
  scale_fill_viridis(discrete = T) +
  scale_color_viridis_d()

graph_ind_3 <- coord_ind |> 
  mutate(rowname = fct_reorder(rowname, Dim.3, .desc = T)) |> 
  ggplot(aes(Dim.3, rowname)) +
  geom_col(aes(colour = df_cuc_rare$clusters, fill = df_cuc_rare$clusters)) +
  theme(legend.position = "inside",
    legend.position.inside = c(1,1),
    legend.justification.inside = c(1,1),
    axis.text = element_text(size = 9, face = "bold"),
    axis.title = element_text(face = "bold")
    ) + 
  xlab(label =  "The sites' positioning in relation to the third principal component") + 
  ylab(label =  "Sites") +
  scale_fill_viridis(discrete = T) +
  scale_color_viridis_d()
```

PCA - bring it home
```{r}
graph_dim1_cors / graph_ind_1 
graph_dim2_cors / graph_ind_2 
graph_dim3_cors / graph_ind_3 
```

# Biplot of Ti vs Zr
The distribution of the elemental ratio of Ti/Zr (titanium/zirconium) can be used to indicate heterogeneity of parent materials because they weather slowly and can retain the characteristics of parent materials (Encyclopedia of Soils in the Environment). 

```{r}
df_pca <- full_join(df_cuc_rare, df_cuc_check_point) 
df_pca$clusters <- coalesce(df_pca$clusters, "outlier")
df_pca$clusters <- as.factor(df_pca$clusters)

ggplot(df_pca, aes(x = Ti_avg, y = Zr_avg, color = clusters)) +
  geom_point(show.legend = T) +
  geom_text(label = df_pca$TIP, hjust = 0.0005, nudge_y = -0.0005, check_overlap = TRUE, show.legend = F) +
  geom_text(label = df_pca$SAMPLE, hjust = 0.0005, nudge_y = 0.0005, nudge_x = -0.0005, check_overlap = TRUE, show.legend = F)
  

```

